---
title: "EWAS Power Calculations (Placenta WGBS)"
author: "George Kuodza"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    number_sections: true
---

# Overview

Power calculations were conducted using the **EpipwR** package under the following assumptions:

- Sample size: **n = 900, 1000, 1100, 1200**
- Tissue: **Placenta**
- dm (true non-null CpGs/regions): **500 and 1000**
- Error control: **FDR = 0.05**
- CpG universe: **20M and 15M**
- Region universe: **500K and 100K**
- Continuous exposure: **correlation-based effect sizes (Pearson’s coefficient) evaluated over a grid of rho_mu values**
- Binary exposure: **25/75 split and 50/50 split**
- Target power: **80%**

---
```{r setup, message=FALSE, warning=FALSE}
library(EpipwR)
library(dplyr)
library(purrr)
library(readr)
library(tibble)
library(kableExtra)

# Inputs
n <- c(900, 1000, 1100, 1200)
Tissue <- "Placenta"
emp_data <- NULL

dm_values <- c(500, 1000)

Totals <- list(
  CpG_20M     = 2e7,
  CpG_15M     = 1.5e7,
  Region_500K = 5e5,
  Region_100K = 1e5,
  Region_20K  = 2e4
)

rho_grid   <- c(0.05, 0.07, 0.10, 0.15, 0.18, 0.20)
delta_grid <- c(0.01, 0.02, 0.03, 0.05)

use_fdr <- TRUE
fdr_fwer <- 0.05

n1_props <- c(`25_75` = 0.25, `50_50` = 0.50)

target_power <- 0.80

out_dir <- "EpipwR_power_rerun_outputs"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

target_power <- 0.80
```

```{r helper}
extract_mdes <- function(df, effect_col, power_col = "avg_power", target_power = 0.80) {

  if (!effect_col %in% names(df)) stop("Missing effect column: ", effect_col)
  if (!power_col %in% names(df)) stop("Missing power column: ", power_col)

  df_sorted <- df %>% arrange(.data[[effect_col]])

  hit <- df_sorted %>%
    filter(.data[[power_col]] >= target_power) %>%
    slice(1)

  if (nrow(hit) == 0) {
    max_row <- df_sorted %>% slice(n())
    tibble(
      target_power = target_power,
      mdes_effect = NA_real_,
      achieved_power_at_mdes = NA_real_,
      max_tested_effect = max_row[[effect_col]],
      achieved_power_at_max = max_row[[power_col]]
    )
  } else {
    tibble(
      target_power = target_power,
      mdes_effect = hit[[effect_col]][1],
      achieved_power_at_mdes = hit[[power_col]][1],
      max_tested_effect = max(df_sorted[[effect_col]]),
      achieved_power_at_max = df_sorted %>% slice(n()) %>% pull(.data[[power_col]])
    )
  }
}
```

```{r run_power, message=TRUE}
# --------------------------
# 1) Region continuous (rho)
# --------------------------
out_cont_Region <- map_dfr(dm_values, function(dm) {
  map_dfr(c("Region_500K", "Region_100K", "Region_20K"), function(total_name) {
    Total <- Totals[[total_name]]

    get_power_cont(
      dm = dm, Total = Total, n = n,
      fdr_fwer = fdr_fwer,
      rho_mu = rho_grid, rho_sd = 0,
      Tissue = Tissue, emp_data = emp_data,
      use_fdr = use_fdr, test = "pearson"
    ) %>%
      mutate(
        Level = "Region",
        Model = "Continuous (rho; Pearson correlation)",
        Total_name = total_name,
        Total_tests = Total,
        dm = dm
      )
  })
})

mdes_cont_Region <- out_cont_Region %>%
  group_by(sample_size, dm, Total_name, Total_tests, Level, Model) %>%
  group_modify(~ extract_mdes(.x, effect_col = "rho_mu", target_power = target_power)) %>%
  ungroup()

# --------------------------
# 2) Region binary (delta)
# --------------------------

out_bin_Region <- map_dfr(dm_values, function(dm) {
  map_dfr(c("Region_500K", "Region_100K", "Region_20K"), function(total_name) {
    Total <- Totals[[total_name]]

    map_dfr(names(n1_props), function(split_name) {
      n1_prop <- n1_props[[split_name]]

      get_power_cc(
        dm = dm, Total = Total, n = n,
        fdr_fwer = fdr_fwer,
        delta_mu = delta_grid, delta_sd = 0,
        n1_prop = n1_prop,
        Tissue = Tissue, emp_data = emp_data,
        use_fdr = use_fdr, test = "pooled"
      ) %>%
        mutate(
          Level = "Region",
          Model = paste0("Binary (delta; ", split_name, ")"),
          Total_name = total_name,
          Total_tests = Total,
          dm = dm,
          n1_prop = n1_prop
        )
    })
  })
})

mdes_bin_Region <- out_bin_Region %>%
  group_by(sample_size, dm, Total_name, Total_tests, Level, Model, n1_prop) %>%
  group_modify(~ extract_mdes(.x, effect_col = "delta_mu", target_power = target_power)) %>%
  ungroup()

# --------------------------
# 3) CpG continuous (rho)
# --------------------------
out_cont_CpG <- map_dfr(dm_values, function(dm) {
  map_dfr(c("CpG_20M", "CpG_15M"), function(total_name) {
    Total <- Totals[[total_name]]

    get_power_cont(
      dm = dm, Total = Total, n = n,
      fdr_fwer = fdr_fwer,
      rho_mu = rho_grid, rho_sd = 0,
      Tissue = Tissue, emp_data = emp_data,
      use_fdr = use_fdr, test = "pearson"
    ) %>%
      mutate(
        Level = "CpG",
        Model = "Continuous (rho; Pearson correlation)",
        Total_name = total_name,
        Total_tests = Total,
        dm = dm
      )
  })
})

mdes_cont_CpG <- out_cont_CpG %>%
  group_by(sample_size, dm, Total_name, Total_tests, Level, Model) %>%
  group_modify(~ extract_mdes(.x, effect_col = "rho_mu", target_power = target_power)) %>%
  ungroup()

# --------------------------
# 4) CpG binary (delta)
# --------------------------
out_bin_CpG <- map_dfr(dm_values, function(dm) {
  map_dfr(c("CpG_20M", "CpG_15M"), function(total_name) {
    Total <- Totals[[total_name]]

    map_dfr(names(n1_props), function(split_name) {
      n1_prop <- n1_props[[split_name]]

      get_power_cc(
        dm = dm, Total = Total, n = n,
        fdr_fwer = fdr_fwer,
        delta_mu = delta_grid, delta_sd = 0,
        n1_prop = n1_prop,
        Tissue = Tissue, emp_data = emp_data,
        use_fdr = use_fdr, test = "pooled"
      ) %>%
        mutate(
          Level = "CpG",
          Model = paste0("Binary (delta; ", split_name, ")"),
          Total_name = total_name,
          Total_tests = Total,
          dm = dm,
          n1_prop = n1_prop
        )
    })
  })
})

mdes_bin_CpG <- out_bin_CpG %>%
  group_by(sample_size, dm, Total_name, Total_tests, Level, Model, n1_prop) %>%
  group_modify(~ extract_mdes(.x, effect_col = "delta_mu", target_power = target_power)) %>%
  ungroup()

full_results <- bind_rows(out_cont_CpG, out_bin_CpG, out_cont_Region, out_bin_Region)

mdes_results <- bind_rows(mdes_cont_CpG, mdes_bin_CpG, mdes_cont_Region, mdes_bin_Region) %>%
  arrange(Level, Total_tests, Model, sample_size, dm)

# Print MDES summary in the knitted doc
mdes_results
```

---

```{r show_mdes_kable, results='asis'}
knitr::kable(
  mdes_results,
  caption = "Minimum Detectable Effect Size (MDES) summary (target power = 0.80)"
)
```

```{r export, message=FALSE}
# Export full tables
write_csv(out_cont_CpG,    file.path(out_dir, "EpipwR_full_CpG_cont.csv"))
write_csv(out_bin_CpG,     file.path(out_dir, "EpipwR_full_CpG_binary.csv"))
write_csv(out_cont_Region, file.path(out_dir, "EpipwR_full_Region_cont.csv"))
write_csv(out_bin_Region,  file.path(out_dir, "EpipwR_full_Region_binary.csv"))

# Export combined tables
write_csv(full_results, file.path(out_dir, "EpipwR_full_results_ALL.csv"))
write_csv(mdes_results, file.path(out_dir, "EpipwR_MDES_summary_ALL.csv"))
```

```{r plot}
plot_dir <- file.path(out_dir, "plots")
if (!dir.exists(plot_dir)) dir.create(plot_dir, recursive = TRUE)

for (tot in unique(out_cont_CpG$Total_name)) {
  for (dm in unique(out_cont_CpG$dm)) {

    df <- out_cont_CpG %>% filter(Total_name == tot, dm == dm)

    p <- EpipwR_plot(df)
    
    print(p)

    ggplot2::ggsave(
      filename = file.path(plot_dir, paste0("CpG_continuous_", tot, "_dm", dm, ".pdf")),
      plot = p,
      width = 8, height = 6, units = "in"
    )
  }
}

for (tot in unique(out_cont_Region$Total_name)) {
  for (dm in unique(out_cont_Region$dm)) {

    df <- out_cont_Region %>% filter(Total_name == tot, dm == dm)

    p <- EpipwR_plot(df)
    
    print(p)

    ggplot2::ggsave(
      filename = file.path(plot_dir, paste0("Region_continuous_", tot, "_dm", dm, ".pdf")),
      plot = p,
      width = 8, height = 6, units = "in"
    )
  }
}

for (tot in unique(out_bin_CpG$Total_name)) {
  for (dm in unique(out_bin_CpG$dm)) {
    for (model in unique(out_bin_CpG$Model)) {

      df <- out_bin_CpG %>% filter(Total_name == tot, dm == dm, Model == model)

      p <- EpipwR_plot(df)
      
      print(p)

      safe_model <- gsub("[^A-Za-z0-9_]+", "_", model)

      ggplot2::ggsave(
        filename = file.path(plot_dir, paste0("CpG_binary_", tot, "_dm", dm, "_", safe_model, ".pdf")),
        plot = p,
        width = 8, height = 6, units = "in"
      )
    }
  }
}


for (tot in unique(out_bin_Region$Total_name)) {
  for (dm in unique(out_bin_Region$dm)) {
    for (model in unique(out_bin_Region$Model)) {

      df <- out_bin_Region %>% filter(Total_name == tot, dm == dm, Model == model)

      p <- EpipwR_plot(df)
      
      print(p)

      safe_model <- gsub("[^A-Za-z0-9_]+", "_", model)

      ggplot2::ggsave(
        filename = file.path(plot_dir, paste0("Region_binary_", tot, "_dm", dm, "_", safe_model, ".pdf")),
        plot = p,
        width = 8, height = 6, units = "in"
      )
    }
  }
}

```
